---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "  "
subtitle: "Web address for GitHub repository"
author: "Camila Zarate Ospina, Savannah Artusi, and Autumn Dunn"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

# Load your packages
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, ggpubr, dplyr, zoo, dataRetrieval)

# Set your ggplot theme
mytheme <- theme_grey() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)
```

Input Raw Data
```{r}
# Load your datasets
#Identify gauge to download
siteNo <- '02081094' 

#Identify parameter of interest: 
pcode = '00065' #gage height (feet)

#Identify statistic code for daily values: 
scode = "00003"  #mean

#Identify start and end dates
start.date = "2001-01-01"
end.date = "2018-12-01"

#Load in data using the package's "readNWISdv" function
Water <- readNWISdv(siteNumbers = siteNo, 
                    parameterCd = pcode, 
                    statCd = scode, 
                    startDate=start.date, 
                    endDate=end.date)

#Rename columns to something more useful
Water <- renameNWISColumns(Water); 
colnames(Water)
saveRDS(Water, file ="../Data/Raw/Water.rds")
```

Process Data
```{r}
Water$Date<-as.Date(Water$Date, format = "%Y/%m/%d")
Water_2001 <- Water %>% 
                  select(Date, GH) %>% 
                  mutate(Month = month(Date),
                         Year = year(Date)) %>% 
                  filter(Month %in% ("9")) %>% 
                  filter(Year %in% ("2001")) %>% 
                  select(Date, GH) %>% 
                  rename(GageHeight = "GH")

Water_2018 <- Water %>% 
                  select(Date, GH) %>% 
                  mutate(Month = month(Date),
                         Year = year(Date)) %>% 
                  filter(Month %in% ("9")) %>% 
                  filter(Year %in% ("2018")) %>% 
                  select(Date, GH) %>% 
                  rename(GageHeight = "GH")
          
Combo_wide <-    bind_cols(Water_2001, Water_2018)   
Combo_long <- full_join(Water_2001, Water_2018, by = c("Date", "GageHeight"))

test<-WaterDate("02081094")
```


# Rationale and Research Questions



\newpage

# Dataset Information



\newpage

# Exploratory Analysis 



\newpage

# Analysis



## Question 1: how much does a stream gage height increase after a hurricane? 

## Question 2: how long until stream gage height returns to normal after a hurricane?




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
